---
title: "Analysis (draft)"
output:
  html_document: 
    toc: yes
    number_sections: false
    toc_collapse: yes
    toc_float: yes
author: "by: Siard van den Bosch"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Analysis (draft)
```{r echo=TRUE, message=FALSE, warning=FALSE}
# Load RSiena package
library(RSiena)
library(sna)
library(network)
```

Generate simulated collaboration network data with 50 nodes (researchers) and 3 time points
```{r echo=TRUE}
set.seed(123)
net1 <- rgraph(50, mode = "graph")
net2 <- rgraph(50, mode = "graph")
net3 <- rgraph(50, mode = "graph")

head(net1, 5)
```

### Simulate 3 networks with 50 nodes each to create longitudinal network data required for RSiena.

The networks must be stored in an array with dimensions nodes x nodes x timepoints. This organizes the data properly for analysis in RSiena.
```{r echo=TRUE}
# Create an array to store the networks
nets <- array(c(net1, net2, net3), dim = c(50, 50, 3))
```

### Convert the array into a RSiena network object
This step is required for further analysis. "oneMode" indicates a single set of nodes.
```{r echo=TRUE}
(net <- sienaNet(nets, type = "oneMode"))
```

### Create a node attribute covariate for gender.
This is required to test gender homophily and differences in network position.
```{r echo=TRUE}
# Generate node attribute for gender (male = 1, female = 2)
(gender <- sample(1:2, 50, replace = TRUE))
```

### Convert the gender node attribute into a RSiena covariate object for use in effects.
```{r echo=TRUE}
# Create gender covariate
(gender <- coCovar(gender))
```

### Combine network and covariates into one RSiena data object required for model estimation.
```{r echo=TRUE}
# Create siena data object
(data <- sienaDataCreate(net, gender))
```


### Generate an initial effetcs object based on the variables in the data.
This specifies the effects to include in the RSiena model.
```{r echo=TRUE}
# Get initial effects
(eff <- getEffects(data))
```


### Add structural effects for transitivity and 3-cycles.
This controls for basic structural network tendencies.
```{r echo=TRUE}
# Include structural effects
(eff <- includeEffects(eff, transTrip, cycle3))
```


### Include homophily effect for the gender covariate.
This will test gender similarity in collaborations (H1).
```{r echo=TRUE}
# Include gender homophily effect
(eff <- includeEffects(eff, simX, interaction1 = gender))
```


Create model object to store results. "colab_model" gives the output file a name.
```{r echo=TRUE}
# Model creation
(mod <- sienaModelCreate(projname = "colab_model"))
```
### Run the model
[Why is `siena07` the most suitable method?](https://siardv.github.io/LabJournal/rsiena_analysis_method.html)
```{r echo=TRUE}
(ans <- readRDS("addfiles/ans.rds"))
# (ans <- siena07(mod, data = data, effects = eff))
```

Estimate model parameters based on data and effects. Returns model results.

```{r echo=TRUE}
# Interpret results
summary(ans)
```

## Test hypotheses
Check significance and direction of effects to assess support for hypotheses.

The positive and significant parameter for the gender homophily effect would indicate support for H1 - researchers collaborate with others of the same gender.

To test the other hypotheses, we would need additional network metrics as covariates, such as:

### Betweenness centrality
```{r echo=TRUE}
# gender_bet <- betweenness(net, v = V(net)$gender)
```

### Convert betweenness scores into a RSiena covariate.
Calculate betweenness centrality scores separately for each gender. This is needed to test if position in the network differs by gender (H2).
```{r echo=TRUE}
# bet <- coCovar(gender_bet)
```

### Ego network density
Calculate ego network density separately for each gender. This will be used to test for differences in local network interconnectedness (H3).
```{r echo=TRUE}
# ego_dens <- ego.density(net1, v = V(net1)$gender)
```


### Create density covariate
Convert the ego density scores into a RSiena covariate.
```{r echo=TRUE}
# dens <- coCovar(ego_dens)
```


### Average path length
Calculate average path length in network. Used to test for differences in global network navigation (H4).
```{r echo=TRUE}
# av_path <- average.path.length(net1, directed = FALSE)
```


### Diameter
Calculate network diameter. Also indicates global network navigability (H4).
```{r echo=TRUE}
# diam <- diameter(net1, directed = FALSE)
```

### Clustering coefficient
Calculate global clustering coefficient. Tests if local network interconnectedness differs by gender (H5).

Include appropriate effects with these covariates to test hypotheses:

```{r echo=TRUE}
# clust <- transitivity(net1, type = "global")
```


**H2**
Include effect of betweenness on incoming ties to test if position affects collaborations.

```{r echo=TRUE}
# eff <- includeEffects(eff, altX, interaction1 = bet)
```


**H3**
Effect of ego density on out ties tests if local density shapes collaborations.

```{r echo=TRUE}
# eff <- includeEffects(eff, egoX, interaction1 = dens)
```


**H4**
Effect of path length on in ties tests for differences in global network reach.

```{r echo=TRUE}
# eff <- includeEffects(eff, altX, interaction1=av_path)
```


**H5**
Effect of diameter on in ties also indicates differences in global network structure.

```{r echo=TRUE}
# eff <- includeEffects(eff, altX, interaction1=diam)
```


**H6**
Ego effect for clustering tests if local cohesion affects collaborations.  

```{r echo=TRUE}
# eff <- includeEffects(eff, egoX, interaction1=clust)
```

