---
title: "Summary of Chapter 7 Methods"
output:
  html_document:
    toc: true
    number_sections: true
    highlight: kate
author: "by: Siard van den Bosch"
bibliography: references.bib
csl: apa.csl
---

# Summary of Chapter 7 Methods 
## Overview

Chapter 7 discusses different methods for analyzing the four theoretical dimensions of social networks (size, composition, structure, evolution) in R. The main methods covered are:

- Exponential-family Random Graph Models (ergm) 
- Stochastic Actor Oriented Models (RSiena)

The chapter recommends trying both methods, but focuses more on RSiena for several reasons:

1. RSiena can model both network structure and composition evolution to distinguish selection and influence processes

2. The author has more expertise with RSiena than ergm

3. This book adds to existing RSiena tutorials

## RSiena Logic

### Ministep assumption
```{r echo=TRUE}
# RSiena ministep logic

# Set up example network
set.seed(123)
(network <- matrix(sample(0:1, size = 25, replace = TRUE), ncol = 5))

# Select random ego
(ego <- sample(1:nrow(network), size = 1))

# Get ego's outgoing ties
out_ties <- network[ego, ]

# Create all options for one tie change
(options <- lapply(1:ncol(network), function(i) {
  temp <- out_ties 
  temp[i] <- 1 - temp[i]
  return(temp)
}))

# Calculate evaluation for each option
(eval <- sapply(options, function(x) sum(x)))

# Pick one option based on evaluation
(probs <- exp(eval) / sum(exp(eval)))
(choice <- sample(1:length(options), size = 1, prob = probs))
(new_out_ties <- options[[choice]])

# Update network with ministep 
(network[ego, ] <- new_out_ties)

# Repeat process for next ministep
(ego <- sample(1:nrow(network), size = 1))
# etc...
```
The key assumption of RSiena models is the **ministep**:

- All network changes are decomposed into very small steps where one actor creates or terminates one outgoing tie
- Does not reflect any particular theory of action, simply an estimation procedure
- Actors change ties to improve satisfaction with local network, evaluating one ministep at a time
- No reactivity, simultaneity, coordination, negotiation, maximization of total utility, altruism, strategic behavior, or investment

### Simulation logic
```{r echo=TRUE}
# Example code to demonstrate simulation logic
set.seed(123)

# 1. Sample ego
(ego <- sample(1:5, size = 1))

# 2. Construct possible networks
(options <- list(matrix(0, 5, 5), matrix(1, 5, 5)))

# 3. Calculate evaluation
(eval <- sapply(options, function(x) sum(x[ego, ])))

# 4. Ego picks network
(choice <- sample(1:length(options), size = 1, prob = exp(eval) / sum(exp(eval))))

# 5. Repeat
for (i in 1:5) {
  ego <- sample(1:5, size = 1)
  options <- list(matrix(0, 5, 5), matrix(1, 5, 5))
  eval <- sapply(options, function(x) sum(x[ego, ]))
  choice <- sample(1:length(options),
    size = 1,
    prob = exp(eval) / sum(exp(eval))
  )
}
```


RSiena models network evolution as an agent-based simulation:

1. Sample ego
2. Construct possible networks after ego ministeps 
3. Calculate ego evaluation of networks
4. Ego picks network (decides on tie change)
5. Repeat until enough ministeps

### Estimation logic 

RSiena estimation:

1. Define model, specify effects
2. Set initial parameter values (often 0)
3. Simulate outcome network 
4. Compare to observed outcome network
5. Update parameters  
6. Repeat until convergence
7. Assess fit on multiple simulations

## RSiena Example

Simple RSiena analysis on co-author network with two waves:

- Define dependent variable (adjacency matrices)
- Specify independent variables (gender) 
- Create RSienaData object
- Define effects in model
- Estimate model with siena07()
- Assess model fit with sienaGOF() on auxiliary statistics

```{r}
# RSiena analysis example
```

## Interpretation of Parameters
```{r echo=TRUE}
# Rate parameter 
rate <- 10

# Network statistic
beta <- 2  
P1 <- 0.5
(P2 <- exp(beta)*P1)
```

- **Rate parameter**: average opportunities for change per actor 
- **Network statistics**: importance of local network configurations for tie changes
- Like logistic regression coefficients: odds ratio = exp(beta)

## Goodness of Fit

- Compare observed vs expected statistics not directly modeled
- Poor fit -> add effects, guided by sienaGOF
- Assess distribution of outdegrees, indegrees, triads, distances, etc
- Violin plots show observed vs simulated
- Small p-value indicates poor fit

```{r eval=FALSE, echo=TRUE}
# Define statistics
tri <- triadcensus(...)
dist <- distanceDistribution(...) 

# GOF 
gof_tri <- sienaGOF(ans, tri, ...)
gof_dist <- sienaGOF(ans, dist, ...)

# Plot
plot(gof_tri)
plot(gof_dist)
```

## Conclusion

- RSiena is a useful method for modeling network evolution as a simulation
- Must understand ministep assumption limitations
- Estimate, simulate, and assess fit for plausible generative model
- Combine with theory and other methods like ERGMs

```{r eval=FALSE, echo=TRUE}
# Conclusion
```