---
title: "Examining Longitudinal Network Data to Determine Waves for RSiena Analysis"
output:
  html_document:
    toc: true
    number_sections: true
    highlight: kate
author: "by: Siard van den Bosch"
bibliography: references.bib
csl: apa.csl
---

# Examining Longitudinal Network Data to Determine Waves for RSiena Analysis
Dividing longitudinal network data into suitable time waves is crucial for modeling network evolution over time using RSiena. This involves balancing detail versus model complexity. This code provides data exploration to guide wave determination based on time points, network changes, visualization, and theory.

```{r eval=FALSE, echo=TRUE}
# Load packages
library(network)
library(tidyverse)
library(gganimate) # for network animation
```

```{r eval=FALSE, echo=TRUE}
#' Generates a random directed network graph with n nodes and m edges
#' Checks that n and m are valid integers >= 0 and that m <= n*(n-1)/2
generate_network <- function(n, m) {
  #' Stops if n or m are invalid
  if (!is.numeric(n) || !is.numeric(m) || n < 0 || m < 0 || round(n) != n || round(m) != m) {
    stop("n and m must be valid nonnegative integers")
  }

  #' Stops if m > n*(n-1)/2
  if (m > n * (n - 1) / 2) {
    stop("m must be <= n*(n-1)/2 for a simple directed graph")
  }

  #' Creates an empty directed graph with n nodes
  g <- igraph::graph.empty(n, directed = TRUE)

  #' Samples 2*m node IDs with replacement to generate m random edges
  edge_pairs <- sample(1:n, size = 2 * m, replace = TRUE)

  #' Arranges the node IDs into an edge list
  edge_pairs <- matrix(edge_pairs, ncol = 2, byrow = TRUE)

  #' Adds the m random edges to the graph
  igraph::add_edges(g, edges = edge_pairs)
}
```


```{r eval=FALSE, echo=TRUE}
# Load longitudinal network data
networks <- readRDS("networks.RDS") # e.g. 50x50 matrix per timepoint  
```

# Calculate time gaps
```{r eval=FALSE, echo=TRUE}
time_points <- seq(min(networks$time), max(networks$time), by = "month")
time_deltas <- diff(time_points)
```

# Calculate % network change between time points
```{r eval=FALSE, echo=TRUE}
net_changes <- purrr::map2_dbl(
  networks[-1], networks[-length(networks)],
  ~ sum(.x != .y) / (nrow(.x)^2)
)
```
# Plot networks 
```{r eval=FALSE, echo=TRUE}
# p <- plot_network(networks, animation = TRUE, layout = "kamadakawai")
# p + animation_save("network_vis.gif")
```

Identify potential wave splits based on:
- Time gaps
- Net changes
- Visual inspection
- Theory/research question

```{r eval=FALSE, echo=TRUE}
wave_splits <- c(3, 7)  
```

# Split data into waves
```{r eval=FALSE, echo=TRUE}
wave_list <- split(networks, cut(seq_along(networks),
  breaks = wave_splits,
  labels = paste0("Wave", 1:length(wave_splits))
))
```

# Print wave summary
```{r eval=FALSE, echo=TRUE}
purrr::map(wave_list, ~ tibble::tbl_df(tibble::tibble(
  wave = .y,
  time_points = purrr::map_int(wave_list[[.y]], "time"),
  mean_change = mean(net_changes[seq_along(.x)])
)))
```