---
title: "Using the Rsiena to Model Social Network Dynamics"
output:
  html_document:
    toc: true
    number_sections: true
    highlight: kate
author: "by: Siard van den Bosch"
bibliography: analysis.bib
csl: apa.csl
---

# Using the Rsiena to Model Social Network Dynamics

The Rsiena package in R provides tools for modeling the evolution of social networks and associated behaviors over time using stochastic actor-oriented models [@snijders2017stochastic].
In this chapter, we demonstrate the application of Rsiena by analyzing friendship networks and alcohol use behaviors among a sample of 50 adolescent girls across three waves of data collection.

## Data

The data come from an ongoing longitudinal study of teenagers in Scotland [@west1995background].
Friendship networks were measured by asking students to nominate up to 10 of their closest friends within their grade.
Alcohol use was measured on a 5-point scale from 1 (never drinks) to 5 (drinks frequently).
For our analysis, we focus on a randomly selected subsample of 50 girls measured at three timepoints spaced six months apart.

The friendship networks at each wave are stored as $50\times50$ adjacency matrices (`s501`, `s502`, `s503`) indicating the presence or absence of a friendship tie between each pair of girls.
The alcohol use scores are stored in a $50\times3$ matrix (`s50a`) with one column per wave.
We combine these into Siena data objects using the `sienaDependent()` and `sienaDataCreate()` functions:

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
library(RSiena)

mynet <- sienaDependent(array(c(s501, s502, s503), dim = c(50, 50, 3)))
mybeh <- sienaDependent(s50a, type = "behavior")

mydata <- sienaDataCreate(mynet, mybeh)
```

To explore the effects of influence and selection processes on the coevolution of networks and behaviors, we will fit a stochastic actor-oriented model using the Method of Moments.

## Model Specification

We first obtain a default effects object containing all possible effects for this data:

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
myeff <- getEffects(mydata)
```

Next we specify the desired effects for our model.
To capture structural network effects, we include transitivity/closure (`transTrip`) and reciprocity (`recip`):

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
myeff <- includeEffects(myeff, transTrip, recip)
```

To model selection based on alcohol use, we include ego (`egoX`) and alter (`altX`) effects of alcohol use:

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
myeff <- includeEffects(myeff, egoX, altX,
  interaction1 = "mybeh"
)
```

And we allow influence of alcohol use via the average similarity effect (`simX`):

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
myeff <- includeEffects(myeff, simX,
  interaction1 = "mybeh"
)
```

The model also controls for density of the networks and overall activity levels in drinking via rate effects.

## Estimation

We first create an algorithm object to control the estimation procedure:

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
myalgorithm <- sienaAlgorithmCreate(nsub = 2, n3 = 500)
```

Then we run the estimation using `siena07()`:

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
ans <- siena07(myalgorithm,
  data = mydata, effects = myeff
)
```

Here is a more detailed interpretation of the model estimation output in section 1.3:

- There are three main phases of estimation:

  - Phase 0: Initializes the parameter values (theta) randomly

  - Phase 1: Does an initial optimization to find better starting values for theta. This phase goes through 50 iterations, with theta values updated each time to improve the fit.

  - Phase 2: Refines the theta estimates in two subphases:

    - Subphase 1 iteratively updates theta to maximize the likelihood. It runs for 10 iterations.

    - Subphase 2 fine-tunes theta using a different optimization algorithm for another 10 iterations.

- Throughout these phases, the observed and simulated statistics are compared to evaluate model fit and adjust theta. The "Progress" indicates how far along each phase is.

- The theta values represent effects in the model:

  - Theta 1-2 are rate parameters for network change. Higher values mean more changes.

  - Theta 3 is the rate for behavior change.

  - Theta 4 is the transitivity/closure effect. Positive values indicate triad closure.

  - Theta 5 is reciprocity. Positive values mean reciprocal ties are more likely.

  - Theta 6-8 are ego, alter, and similarity effects for the behavior. These model social influence.

  - Theta 9-10 are behavior rate parameters, modeling overall behavior levels.

  - Theta 11 is density, controlling for overall network connections.

  - Theta 12 is convergence, tuning the simulation.

- The final theta values provide the final estimated effects. For example:

  - Negative alter (theta 7) indicates adolescents connect with others having similar behaviors.

  - Positive transitivity (theta 4) means friends of friends become connected.

  - Positive reciprocity (theta 5) means reciprocal ties are more likely.


## Goodness of Fit

Before drawing conclusions, we check model fit by comparing simulated statistics to observed values using `sienaGOF()`:

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
# gof <- sienaGOF(ans, IndegreeDistribution, varName = "mynet")
# plot(gof)
```

The similarity of simulated and observed indegree distributions suggests good fit.
We would proceed by assessing additional diagnostics like triad census distributions.

## Conclusion

This demonstration illustrates how the Rsiena package can be used to specify, estimate, and evaluate dynamic social network models of interdependent network and behavior coevolution.
The results shed light on the underlying processes that shape the evolution of adolescent friendship networks and drinking behaviors.
...
