---
title: "Webscraping"
output:
  html_document: 
    toc: yes
    number_sections: false
    toc_collapse: yes
    toc_float: yes
author: "by: Siard van den Bosch"
bibliography: references.bib
csl: apa.csl
---

# Webscraping
## Data Collection

- Used `rvest` package to webscrape HTML table from department website containing faculty names, titles, and contact info
- Queried Crossref API using `rcrossref` package to retrieve metadata for publications by each faculty member 
  - Passed faculty names as search queries
  - Retrieved full bibliographic data including titles, journals, dates, co-authors
  - Handled API errors and missing author results
- Filtered publications to only journal articles for analysis

## Data Cleaning

- Parsed HTML table into a dataframe and cleaned column names
- Added last name column by separating first and last names  
- Standardized faculty names by removing non-word characters
- Used the `gender` R package with 5 different prediction methods (ssa, ipums, napp, kantrowitz, genderize) to predict gender from first names, keeping most confident result
- Joined predicted gender back to faculty data frame
- Unnested article authors into separate rows for each faculty author
- Completed partial dates and formatted dates into YYYY-MM-DD format
- Combined multiple affiliation string columns into single column
- Removed rows and columns with high proportion of missing values

## Resulting Dataset

- Faculty data combined with predicted gender and matched publication metadata
- One row for each faculty author on each published journal article  
- Columns for standard author name, predicted binary gender, affiliation, publication date
- Enables analysis of publishing trends by gender over time

```{r include=FALSE}
stopifnot(gsub(path.expand("~/"), "", getwd()) == "Documents/GitHub/LabJournal")
path_to_publications <- "addfiles/publications.rds"
```

```{r echo=TRUE, message=TRUE, warning=TRUE}
# Whether to pull data from original sources
# When TRUE, fetch live data from remote sources
# When FALSE, use cached/local data
pull_sources <- FALSE

# Load required R packages
load_packages <- function(libs) {
  sapply(setdiff(libs, installed.packages()),
    install.packages,
    dependencies = TRUE
  )
  for (l in libs) {
    library(l, character.only = TRUE)
  }
}

load_packages(c(
  "rvest", "gender", "tibble", "dplyr", "stringr",
  "lubridate", "rcrossref", "igraph"
))

as_data_table <- function(...) {
  stopifnot(is.data.frame(...))
  DT::datatable(...,
    rownames = FALSE,
    options = list(
      searching = FALSE, scrollX = TRUE,
      scrollY = TRUE, pageLength = 20,
      autoWidth = FALSE
    )
  )
}
```

# Retrieve staff data by parsing HTML table from department website
```{r echo=TRUE, message=TRUE, warning=TRUE}
staff_url <- "https://www.ru.nl/sociology/research/staff/"
staff_df <- init <- rvest::read_html(staff_url) %>%
  rvest::html_table(header = TRUE) %>%
  `[[`(1) %>%
  setNames(gsub("\\W+", "", names(.)) %>% tolower())

as_data_table(staff_df)
```

# Adds "family" column with last names only to staff data frame
```{r echo=TRUE, message=TRUE, warning=TRUE}
staff_df$family <- sapply(strsplit(staff_df$staff, ","), function(x) {
  parts <- unlist(strsplit(base::trimws(x), " "))
  if (tolower(tail(parts, 1)) == "van") {
    paste(tail(parts, 1), head(parts, 1))
  } else {
    head(parts, 1)
  }
})

as_data_table(staff_df)
```

# Returns staff data frame with cleaned names
# Add full name column by combining given and family name
```{r echo=TRUE, message=TRUE, warning=TRUE}
staff_df <- staff_df %>%
  transform(staff = gsub("(.*)\\\r.*", "\\1", staff)) %>%
  transform(staff = gsub("\\s+", " ", staff)) %>%
  transform(given = gsub(".*\\((.*)\\).*", "\\1", staff)) %>%
  transform(name = paste(given, family))

as_data_table(staff_df)
```

# Get publications for each staff member by querying Crossref API
```{r echo=TRUE, message=TRUE, warning=TRUE}
get_author_publications <- function(author_name) {
  q <- gsub(" ", "+", tolower(author_name))
  tryCatch(
    rcrossref::cr_works(query = q)$data,
    error = function(e) {
      if (grepl("No author found", e$message)) {
        warning(paste0("No publications for '", author_name, "'"))
      } else {
        warning("Error: ", e$message)
      }
    }
  )
}
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
if (pull_sources) {
  publications <- lapply(staff_df$full_name, get_author_publications)
  saveRDS(publications, path_to_publications)
} else {
  publications <- readRDS(path_to_publications)
}

publications
# contains nested tibble in column 'authors'
```

# Bind publication results and unnest author column lists into separate rows
```{r echo=TRUE, message=TRUE, warning=TRUE}
all_publications <- dplyr::bind_rows(publications) %>%
  tidyr::unnest(cols = author)

all_publications <- all_publications[all_publications$type == "journal-article", ] %>%
  dplyr::select(-abstract) # remove abstract to speed up script

as_data_table(all_publications)
```

# Clean up data
## Cleaning date values through coalescing
```{r echo=TRUE, message=TRUE, warning=TRUE}
all_dates <- all_publications %>%
  dplyr::select(doi, published.print, subject) %>%
  dplyr::distinct() %>%
  `$`("published.print")

all_dates
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
tbl <- table(is.na(all_dates))
missing_str <- sprintf(
  "%.2f%% missing (%d missing, %d non-missing)",
  100 * tbl["TRUE"] / sum(tbl), tbl["TRUE"], tbl["FALSE"]
)
cat(missing_str, "\n")
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
complete_date <- function(x) {
  if (is.na(x) || !grepl("[0-9]{4}", x)) {
    return(NA)
  }
  date_parts <- strsplit(x, "-", fixed = TRUE)[[1]]
  paste0(c(date_parts, rep("01", 3 - length(date_parts))), collapse = "-")
}

format_date <- function(x) {
  if (!grepl("[0-9]{4}", x)) {
    return(x)
  } else if (grepl("^\\d{4}-\\d{2}-\\d{2}$", x)) {
    return(x)
  } else {
    date <- tryCatch(
      lubridate::dmy(x),
      warning = function(w) {
        return(NA)
      },
      error = function(e) {
        return(NA)
      }
    )
    if (is.na(date)) {
      return(x)
    } else {
      return(complete_date(format(date, "%Y-%m-%d")))
    }
  }
}
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
apwcd <- all_publications_with_complete_date <- transform(all_publications,
  published =
    ifelse(!is.na(published.print) & !is.na(published.online),
      ifelse(nchar(published.print) > nchar(published.online),
        published.print, published.online
      ),
      dplyr::coalesce(published.print, published.online)
    )
)

as_data_table(apwcd)
```

```{r echo=TRUE, message=TRUE, warning=TRUE}
dates_from_assertion <-
  apwcd %>%
  dplyr::select(doi, published, assertion) %>%
  dplyr::distinct() %>%
  dplyr::filter(length(assertion) > 0) %>%
  tidyr::unnest(cols = c(assertion)) %>%
  dplyr::filter(grepl("first_online|published", name)) %>%
  transform(published = sapply(value, format_date))

apwcd <- dplyr::bind_rows(
  apwcd[!apwcd$doi %in% dates_from_assertion$doi, ],
  dates_from_assertion
) %>%
  dplyr::arrange(doi)

as_data_table(apwcd)
```


```{r}
# Checks is list-column
is_list_column <- function(x) all(apply(x, 1, is.recursive), na.rm = TRUE)


```

```{r echo=TRUE, message=TRUE, warning=TRUE}
apwcd$published <- sapply(apwcd$published, complete_date, USE.NAMES = FALSE)

affiliation_columns <- grep("affiliation", names(apwcd), value = TRUE)
apwcd <- apwcd %>%
  dplyr::mutate(affiliation = dplyr::coalesce(!!!rlang::syms(affiliation_columns)))
apwcd <- apwcd[!names(apwcd) %in% affiliation_columns]

apwcd <- apwcd[!sapply(apwcd, is.list)]
apwcd <- apwcd[, !(colSums(is.na(apwcd)) / nrow(apwcd)) > 0.20]
apwcd <- apwcd[!c(rowSums(is.na(apwcd)) > ncol(apwcd) / 2), ]

as_data_table(apwcd)
```


# Test all gender prediction methods
```{r echo=TRUE, message=TRUE, warning=TRUE}
gender_prediction_methods <-
  c("ssa", "ipums", "napp", "kantrowitz", "genderize")
all_gender_predictions <-
  lapply(gender_prediction_methods, function(method) {
    gender::gender(staff_df$given, method = method)
  })

all_gender_predictions
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
determine_gender_by_multiple_names <- function(name_array, method) {
  if (is.null(name_array) || length(name_array) == 0) {
    return(NA)
  }
  pred <- lapply(name_array, gender::gender, method = method)
  pred_df <- do.call(rbind, pred)

  if (!is.null(dim(pred_df)) && nrow(pred_df) == 0) {
    return(NA)
  } else if (!is.null(dim(pred_df)) && nrow(pred_df) > 1) {
    props <- colMeans(pred_df[c("proportion_male", "proportion_female")])
    data.frame(
      name = paste(name_array, collapse = " "),
      rbind(props),
      gender = which.max(props) %>% names() %>% gsub(pattern = "proportion_", replacement = ""),
      row.names = NULL
    )
  } else {
    return(pred[[1]])
  }
}
```


# Predict gender using "ipums" then "genderize" if no match
```{r echo=TRUE, message=TRUE, warning=TRUE}
predict_gender <- function(name) {
  name <- gsub("([a-z])([A-Z])", "\\1 \\2", name)
  name_enc <- gsub("\\W", "", iconv(name, to = "ascii//TRANSLIT")) %>%
    stringr::str_to_title()
  pred <- determine_gender_by_multiple_names(name_enc, method = "ipums")
  if (is.null(dim(pred)) || nrow(pred) == 0) {
    pred <- determine_gender_by_multiple_names(name_enc, method = "genderize")
  }
  pred$name <- paste(name_enc, collapse = " ")
  return(pred)
}
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
clean_given_name <- function(given_name) {
  gsub("\\.", " ", given_name) %>%
    stringr::str_squish() %>%
    strsplit(" ") %>%
    `[[`(1) %>%
    `[`(nchar(.) > 1)
}
```


```{r}
get_first_paper_per_affiliation <- function(given, family){
  
}
```

```{r echo=TRUE, message=TRUE, warning=TRUE}
apwcdg <- apwcd_with_gender <-
  apwcd$given %>%
  unique() %>%
  sapply(clean_given_name) %>%
  Filter(f = length) %>%
  lapply(predict_gender) %>%
  dplyr::bind_rows() %>%
  dplyr::rename(given = name) %>%
  dplyr::distinct() %>%
  dplyr::left_join(apwcd, by = "given")

as_data_table(apwcdg)
```


```{r echo=TRUE, message=TRUE, warning=TRUE}
apwcdg <- apwcdg[!is.na(apwcdg$doi), ]

apwcdg %>%
  dplyr::distinct() %>%
  transform(name = paste(given, family)) %>%
  dplyr::select(
    -proportion_male, -proportion_female, -year_min, -year_max,
    -reference.count, -references.count,
    -indexed, -created, -deposited, -issued,
    -issn, -prefix, -url, -source,
    -given, -family
  ) %>%
  dplyr::rename(
    journal = container.title,
    referenced = is.referenced.by.count
  ) -> apwcdg2

as_data_table(apwcdg)
```